A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-23 text/informal-overview.tex

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{stage
  variable}\footnote{%
  In Hanada and Igarash~\cite{Hanada2014}, it was called a
  \textit{transition variable}, which is derived from correspondence
  to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called \textit{stage}.  Intuitively,
they represent the depth of nested brackets.  Stage variables can be
abstracted by $\Lambda\alpha.M$ and instantiated by an application
$M\ A$ to stages.  For example,
$\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10) 5)$ is a code value,
where \(\alpha\) is abstracted.  If it is applied to \(\varepsilon\),
which is the empty sequence standing for zero, \(\TB_\alpha\) becomes
\(\TB_\varepsilon\), which stands for zero brackets, and so, in fact,
\(\TB_\alpha\) disappears.  As a result, the application reduces to
(unquoted) \((\lambda x:\I.x+10) 5\) and to 15.  So, application of a
\(\Lambda\)-abstraction to $\varepsilon$ corresponds to \texttt{run}.
If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
\(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).
These are expressed by the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates nesting of
\(\TB\) and \(\TBL\) (and also \(\%_\alpha\) as we see later).

@@

First, type-level constants such as \(\I\) and Vector can be used at
every stage in \LMD.  Technically, we introduce a signature that
declares kinds of type-level constants and types of constants.  For
example, a signature for the Boolean type and constans is given as
follows $\B::*, \text{true}:\B, \text{false}:\B$ (where $*$ is the
kind of proper types).  Declarations in a siganature are not
associated to particular stages and so they are used at every stage.

>>>>

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{stage
  variable}\footnote{%
  In Hanada and Igarashi~\cite{Hanada2014}, it was called a
  \textit{transition variable}, which is derived from correspondence
  to modal logic, studied by Tsukada and Igarashi~\cite{Tsukada}.} and
a sequence of stage variables is called \textit{stage}.  Intuitively,
they represent the depth of nested brackets.  Stage variables can be
abstracted by $\Lambda\alpha.M$ and instantiated by an application
$M\ A$ to stages.  For example,
$\Lambda\alpha.\TB_\alpha ((\lambda x:\I.x+10) 5)$ is a code value,
where \(\alpha\) is abstracted.  If it is applied to \(\varepsilon\),
which is the empty sequence standing for zero, \(\TB_\alpha\) becomes
\(\TB_\varepsilon\), which stands for zero brackets, and so, in fact,
\(\TB_\alpha\) disappears.  As a result, the application reduces to
(unquoted) \((\lambda x:\I.x+10) 5\) and to 15.  So, application of a
\(\Lambda\)-abstraction to $\varepsilon$ corresponds to \texttt{run}.
If \(\alpha\) is instantiated with, say, \(\beta\gamma\), then
\(\TB_\alpha\) becomes nested brackets \(\TB_{\beta} \TB_\gamma\).
These are expressed by the following reduction rule:
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
where stage substitution \([\alpha \mapsto A]\) manipulates nesting of
\(\TB\) and \(\TBL\) (and also \(\%_\alpha\) as we see later).

@@

First, type-level constants such as \(\I\) and Vector can be used at
every stage in \LMD.  Technically, we introduce a signature that
declares kinds of type-level constants and types of constants.  For
example, a signature for the Boolean type and constants is given as
follows $\B::*, \text{true}:\B, \text{false}:\B$ (where $*$ is the
kind of proper types).  Declarations in a signature are not
associated to particular stages and so they are used at every stage.

### 2019-06-23 text/properties.tex
Inversion for $\Lambda$ type

\begin{lemma}[Inversion]\ 
  \begin{enumerate}
  \item If $\G \V (\lambda x:\sigma.M) : (\Pi x:\sigma'.\tau)@A$,
    then $\G \V \sigma \E \sigma'@A$ and $\G ,x:\sigma@A\V M:\tau @A$.
  \item If $\G \V \TB_\alpha M : \TW_\alpha \tau @A$ then $\G \V M : \tau @A$.
  \item If $\G \V \Lambda\alpha.M : \forall\alpha.\tau @A$ then $\G \V M : \tau @A$ and $\alpha \notin \FTV(\G) \cup \FV(A)$.
  \end{enumerate}
\end{lemma}

>>>>

\begin{lemma}[Inversion]\ 
	\begin{enumerate}
		\item If $\G \V (\lambda x:\sigma.M) : \rho$ then there are $\sigma'$ and $\tau'$ such that
		      $\rho = \Pi x:\sigma'.\tau'$, $\G \V \sigma \E \sigma'@A$ and $\G ,x:\sigma'@A\V M:\tau'@A$.
		\item If $\G \V \TB_\alpha M : \tau@A$ then 
		      there is $\sigma$ such that $\tau = \TW_\alpha \sigma$ and $\G \V M : \sigma@A$.
		  \item If $\G \V \Lambda\alpha.M : \tau$ then 
		  there is $\sigma$ such that $\sigma = \forall\alpha.\sigma$ and $\G \V M : \sigma@A$.% and $\alpha \notin \FTV(\G) \cup \FV(A)$.
	\end{enumerate}
\end{lemma}

### 2019-06-23 text/related-work.tex

\section{Related Work \label{sec:related-work}}

@@

Theoretical studies on multi-stage programming owe a lot to seminal
work by Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, which found Curry-Howard
correspondence between multi-stage calculi and modal logic.  In
particular, Davies' $\lambda^\circ$~\cite{davies1996temporal} has been
a basis for multi-stage calculi with quasi-quotation.  $\lambda^\circ$
did not have operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP}
enhanced and improved $\lambda^\circ$ towards the development of a
type-safe multi-stage calculus with quasi-quotation, run, and CSP,
which were proposed by Taha and Sheard as constructs for multi-stage
programming~\cite{MetaML}.
% Benaissa et al.~\cite{benaissa1999logical} study the relationship between a multi-stage calculus and category theory or modal logic.
% Taha and Sheard introduced run and CSP to a multi-stage calculus in 
Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a
type sound manner and formed a basis of earlier versions of MetaOCaml.
A different approach to type-safe multi-stage programming with
different semantics for quasi-quotations has been studied by Kim, Yi,
and Calcagno~\cite{DBLP:conf/popl/KimYC06}.

>>>>

\section{Related Work}
\label{sec:related-work}

@@

Theoretical studies on multi-stage programming owe a lot to seminal
work by Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, which found Curry-Howard
correspondence between multi-stage calculi and modal logic.  In
particular, Davies' $\lambda^\circ$~\cite{davies1996temporal} has been
a basis for multi-stage calculi with quasi-quotation.  $\lambda^\circ$
did not have operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP}
enhanced and improved $\lambda^\circ$ towards the development of a
type-safe multi-stage calculus with quasi-quotation, run, and CSP,
which were proposed by Taha and Sheard as constructs for multi-stage
programming~\cite{MetaML}.
% Benaissa et al.~\cite{benaissa1999logical} study the relationship between a multi-stage calculus and category theory or modal logic.
% Taha and Sheard introduced run and CSP to a multi-stage calculus in 
Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a
type sound manner and formed a basis of earlier versions of MetaOCaml.
Different approaches to type-safe multi-stage programming with
slightly different constructs for composing and running code values have been studied by Kim, Yi,
and Calcagno~\cite{DBLP:conf/popl/KimYC06} and Nanevski and Pfenning~\cite{DBLP:journals/jfp/NanevskiP05}.
